Radiocarbon measurement of CO~2~ from carbonate minerals using a gas ion source with an open split interface

### 2022-03-25 doc/AMS-15/AMS-15_paper/AMS-15_paper.Rmd
Many edits to text.

Radiocarbon Accelerator Mass Spectrometry (AMS) is a sensitive method for determining the age, provenance, and composition of carbon-containing materials. A primary disadvantage of the method is its cost, which often limits the number of samples analyzed. This forces researchers to carefully choose samples to be run and limits repeat measurements for error determination. One driver of the cost of AMS analysis is the time and effort required for sample preparation, including reduction of the sample to graphite. Analysis of analyte CO2 directly via a gas-accepting AMS ion source eliminates the cost and effort of graphitization and returns results more rapidly. The NOSAMS USAMS is an AMS system based on a 3MV Tandetron accelerator. A NEC 134-sample gas-capable MCSNICS source. Analysis of samples as CO2 simplifies sample preparation, making analysis less expensive and faster.

@@

An open inlet has the advantage of simplicity; flow of CO2 to the source is preset and there are no separate sample transfer and dilution steps. This may translate to greater throughput by eliminating the time required for sample transfer, dilution, and setting of delivery pressure. A disadvantage of the open split is that the flow of CO2 to the source varies with sample size (CO2:He) and changes over the course of a measurement due to dilution of the sample by displacement helium. Our method compensates for this by limiting CO2 flow to a range that produces relatively constant C- ion current and stable ratios. Salazar and Szidat, Roberts gasbench, other open split methods.

The gas-capable 134-sample MCSNICS from National Electrostatics Corporation (NEC) works on the same principal as hybrid sources developed by ETH (REF) and HVEE (REF). Sample CO2 is introduced onto a titanium frit in an aluminum target using stainless tubing mounted to an arm with a probe that mates with the back of the target. A CO~2~ - helium mixture is exposed to a beam of charged cesium atoms, which causes dissociation of CO~2~ and ionization of molecular and atomic species. Negative ions are extracted from the target and a magnetic filter picks species with masses corresponding to carbon.

@@

The flow rate from the open split into the source is defined by 250cm of 50um glass capillary (confirm this, SGE, Trajan, Agilent?), split into two sections by a zero-volume bulkhead connector at the source cage. The section inside the cage is jacketed in teflon tubing to help protect against spark damage.

@@

```{r}
carbcur <- carb %>%  
  map_dfr(2) %>% 
  compare_consensus() %>% 
  mutate(c12ua = he12C * 1E6)
samples <- read_csv(here("data/carb_data.csv")) %>% 
  select(wheel, pos, mass, he_added)
currents <- left_join(carbcur, samples, by = c("wheel", "pos")) %>% 
  filter(wheel != "USAMS020521")

@@

Our initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO2 connected through a regulator to the ion source to measure source performance over a range of CO2 flows. Optimal flow for this source is 7.5uLs-1 CO2, but the source produces near optimal currents over a wide range of flows. Adding helium to the gas mixture produced similar currents to pure CO2 at the same CO2 flow rates.

@@

Precision taken as the standard error of the mean of repeated runs of a sample is around 1%. Precision is typically a little worse than predicted by counting statistics. More work is needed to look at whether this reflects variability between samples, but initial test of paired replicates show similar standard deviation for samples that show good stable currents.

@@

Discuss delivery rates and sample sizes here or in results?

Initial currents of \> 10 uA dropped rapidly, eventually settling to 100-400 nA.

@@

First tests of displacing pure CO2 from a vial into the open split using helium are promising. Helium flowing at 100uLm-1 produces current from 7mL of CO2 for over 2 hours. Currents do decrease with time due to dilution, but by supplying higher than optimal CO2 flow to the source initially, stable currents can be maintained for some time. Fraction of CO2 in sample gas should follow this relationship:

@@

Measurement of radiocarbon-free reference materials shows that the average normalized ratio for these samples is `r blankmean` ± `r blanksd` (N=`r nblank`). We investigated the sources of this method blank as a current-dependent blank and as a current-independent blank.

To assess blank contribution and ratio dependence on sample current, vials of live and dead standard gas were measured as the displacement helium flow diluted the samples, producing diminishing current over time. We used these data to develop a current-balance model following methods used to assess mass-dependent or constant contamination blanks in graphite targets (REFS). The constant contamination blank was also assessed directly by measuring pure helium in vials using the same method. Results from both tests are shown in figure 2. The inverse of the sample current is used to linearize the data, allowing assessment with a simple linear model. Fits from modern and dead dilution tests should trend towards the fraction modern of the blank, crossing at the inverse of blank current. Based on this assessment, we estimate the ratio of the blank as having F14C of 0.43 with a current of 110 nA.

>>>>

Radiocarbon Accelerator Mass Spectrometry (AMS) is a sensitive method for determining the age, provenance, and composition of carbon-containing materials. A primary disadvantage of the method is its cost, which often limits the number of samples analyzed. This forces researchers to carefully choose samples to be run and limits repeat measurements for error determination. One driver of the cost of AMS analysis is the time and effort required for sample preparation, including reduction of the sample to graphite. Analysis of CO2 directly via a gas-accepting AMS ion source eliminates the cost and effort of graphitization and returns results more rapidly. The NOSAMS USAMS is an AMS system based on a 3MV Tandetron accelerator (REF) combined with A NEC 134-sample gas-capable MCSNICS source \@ref(Longworth 2015).

@@

An open inlet has the advantage of simplicity; flow of CO2 to the source is preset and there are no separate sample transfer and dilution steps. This may translate to greater throughput by eliminating the time required for sample transfer, dilution, and setting of delivery pressure. A disadvantage of the open split is that the flow of CO2 to the source varies with sample size (CO2:He) and changes over the course of a measurement due to dilution of the sample by displacement helium. Our method compensates for this by limiting CO2 flow to a range that produces relatively constant C- ion current and stable ratios. In gas-source AMS, Roberts et. al (2013?) use an open split to introduce sample gas to a ECR ion source and Salazar and Szidat (2021?) designed a continuous flow system that maintains relatively stable pCO2 during acquisition by controlling pressure and venting excess gas to atmosphere.

The gas-capable 134-sample MCSNICS from National Electrostatics Corporation (NEC) works on the same principle as hybrid sources developed by ETH (REF) and HVEE (REF). The sample CO~2~ - helium mixture is introduced onto a titanium frit in an aluminum target cathode using stainless tubing mounted to an arm with a probe that mates with the back of the cathode. Sample gas is exposed to a beam of charged cesium atoms, which causes dissociation of CO~2~ and ionization of molecular and atomic species. Negative ions are extracted from the target and carbon species are selected and interferences filtered by the USAMS Tandetron beamline.

@@

The helium displacement flow rate was set to 250ul/min for initial tests and was reduced to 100 ul/min for later work including most reference standard comparisons and the core comparison. The flow rate from the open split into the source is defined by 250cm of 50um glass capillary (confirm this, SGE, Trajan, Agilent?), split into two sections by a zero-volume bulkhead connector at the source cage. The section inside the cage is jacketed in teflon tubing to help protect against spark damage. The total flow rate of gas to the source varies due to the viscosity difference between CO2 and helium and is estimated to be 30 ul/min. ???

@@

```{r, message = FALSE}
carbcur <- carb %>%  
  map_dfr(2) %>% 
  compare_consensus() %>% 
  mutate(c12ua = he12C * 1E6)
samples <- read_csv(here("data/carb_data.csv")) %>% 
  select(wheel, pos, mass, he_added)
currents <- left_join(carbcur, samples, by = c("wheel", "pos")) %>% 
  filter(wheel != "USAMS020521")

@@

Initial work with the NEC source established optimal source parameters for consistent source performance. We used tank CO2 connected through a regulator to the ion source to measure source performance over a range of CO2 flows. Optimal flow for this source is 7.5uLs-1 CO2, but the source produces near optimal currents over a wide range of flows. Adding helium to the gas mixture produced similar currents to pure CO2 at the same CO2 flow rates.

@@

@@

@@

First tests of displacing pure CO2 from a vial into the open split using helium are promising. Helium flowing at 100uLm-1 produces current from 7mL of CO2 for over 2 hours. Currents do decrease with time due to dilution, but by supplying higher than optimal CO2 flow to the source initially, stable currents can be maintained for some time.

Fraction of CO2 in sample gas should follow this relationship:

@@

The HGIS method produces higher ratios for radiocarbon dead samples than the graphite method. Measurement of radiocarbon-free reference materials shows that the average normalized ratio for these samples is `r blankmean` ± `r blanksd` (N=`r nblank`). We investigated the sources of this method blank as a current-dependent blank and as a current-independent blank.

HGIS cathodes produce measurable ion current, even with no gas flow. This current is initially > 10 uA and drops exponentially, reaching a baseline of 100-200 nA after 5-10 minutes. For the range of currents in this study (3-9 uA), this baseline current represents a 1-6% contribution to the total current at the F14C of the target baseline. This can be interpreted as a constant current dependent blank, analogous to a mass dependent "constant contaminant" blank and may be modelled with the same tools (REF). To assess blank contribution and ratio dependence on sample current, vials of live and dead standard gas were measured as the displacement helium flow diluted the samples, producing diminishing current over time. We used these data to develop a current-balance model following methods used to assess mass-dependent or constant contamination blanks in graphite targets (REFS). The constant contamination blank was also assessed directly by measuring pure helium in vials using the same method. Results from both tests are shown in figure 2. The inverse of the sample current is used to linearize the data, allowing assessment with a simple linear model. Fits from modern and dead dilution tests should trend towards the fraction modern of the blank, crossing at the inverse of blank current. Based on this assessment, we estimate the ratio of the blank as having F14C of 0.43 with a current of 110 nA.
